---
title: "Introduction to HiCExperiment"
author: "Jacques Serizay"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{Introduction to HiCExperiment}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, eval = TRUE, echo=FALSE, results="hide", warning=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>", 
    crop = NULL
)
suppressPackageStartupMessages({
    library(dplyr)
    library(GenomicRanges)
    library(HiContactsData)
    library(HiCExperiment)
})
```

# The `HiCExperiment` class

`HiCExperiment` package implements the new `HiCExperiment` S4 class. It is build 
on pre-existing Bioconductor classes, namely `InteractionSet`, 
`GenomicInterations` and `ContactMatrix` 
(`Lun, Perry & Ing-Simmons, F1000Research 2016`), and leverages them to 
import locally stored `.(m)cool` files. It further provides **analytical** 
and **visualization** tools to investigate contact maps directly in `R`. 

```{r}
library(HiCExperiment)
showClass("HiCExperiment")
hic <- contacts_yeast()
hic
```

```{r, eval = TRUE, echo=FALSE, out.width='100%'}
knitr::include_graphics(
   "https://raw.githubusercontent.com/js2264/HiCExperiment/master/man/figures/HiCExperiment_data-structure.png"
)
```

# Citing HiCExperiment 

```{r}
citation('HiCExperiment')
```

# Basics: importing `.(m)cool` or `.hic` files as `HiCExperiment` objects

The `import()` function allows one to import .(m)cool or .hic files in R as 
`HiCExperiment` objects.

```{r, eval = FALSE}
## Not run:
hic <- import("path-to-(m)cool-file", focus = "...", resolution = ...)
## End(Not run)
```

# Import arguments

The `HiContactsData` package gives access to a range of toy datasets stored 
by Bioconductor in the `ExperimentHub`. 

```{r}
library(HiContactsData)
cool_file <- HiContactsData('yeast_wt', format = 'cool')
cf <- CoolFile(cool_file)
cf
import(cf)
```

The `focus` argument is used to specifically import contacts within a genomic 
locus of interest. 

```{r}
range <- 'I:20000-80000' # range of interest
hic <- import(cf, focus = range)
hic
focus(hic)
```

`import()` works with `.mcool` files as well: in this case, the user can  
specify the resolution at which count values are recovered. 

```{r}
mcool_file <- HiContactsData('yeast_wt', format = 'mcool')
range <- 'II:0-800000' # range of interest
lsCoolResolutions(mcool_file, verbose = TRUE)
hic <- import(mcool_file, format = 'cool', focus = range, resolution = 2000)
hic
```

One can also extract a count matrix from a `.(m)cool` file that is *not* 
centered at the diagonal. To do this, specify a couple of coordinates in the 
`focus` argument using a character string formatted as `"... x ..."`: 

```{r}
hic <- import(mcool_file, format = 'cool', focus = 'II:0-500000 x II:100000-600000', resolution = 2000)
focus(hic)
```

## Slots

Slots for a `HiCExperiment` object can be accessed using the following `getters`: 

```{r}
fileName(hic)
focus(hic)
resolutions(hic)
resolution(hic)
interactions(hic)
scores(hic)
tail(scores(hic, 1))
tail(scores(hic, 'balanced'))
topologicalFeatures(hic)
pairsFile(hic)
metadata(hic)
```

Several extra functions are available as well: 

```{r}
seqinfo(hic) ## To recover the `Seqinfo` object from the `.(m)cool` file
bins(hic) ## To bin the genome at the current resolution
regions(hic) ## To extract unique regions of the contact matrix
anchors(hic) ## To extract "first" and "second" anchors for each interaction
```

## Slot setters

### Scores 

Add any `scores` metric using a numerical vector. 

```{r}
scores(hic, 'random') <- runif(length(hic))
scores(hic)
tail(scores(hic, 'random'))
```

### Features 

Add `topologicalFeatures` using `GRanges` or `Pairs`. 

```{r}
topologicalFeatures(hic, 'viewpoints') <- GRanges("II:300000-320000")
topologicalFeatures(hic)
topologicalFeatures(hic, 'viewpoints')
```

## Coercing `HiCExperiment`

Using the `as()` function, `HiCExperiment` can be coerced in `GInteractions`, 
`ContactMatrix` and `matrix` seamlessly.

```{r}
as(hic, "GInteractions")
as(hic, "ContactMatrix")
as(hic, "matrix")[1:10, 1:10]
as(hic, "data.frame")[1:10, ]
```

# Session info

```{r}
sessionInfo()
```
